Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-04-11 manuscript_rmarkdown/manuscript.Rmd
Fixed spelling, grammar and style issues.

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.5, warning=FALSE}
cap <- paste0(
  "Effect of location, scale and outliers on estimation of the Box-Cox and Yeo-Johnson transformation parameter $\\lambda$. ",
  "$10000$ samples were drawn from a normal distribution: ",
  "$\\mathcal{N}(\\mu, 1)$ for the shift dataset, $\\mathcal{N}(10, \\sigma)$ for the scale dataset and $\\mathcal{N}(0, 1)$ for the outlier dataset. ",
  "Additionally, an outlier with value $d$ was added to the outlier dataset. ",
  "Since samples are drawn from a normal distribution, a transformation parameter of $\\lambda = 1$ is expected. ",
  "However, a large shift in location, a scale that is small compared to the location, or presence of large outliers ",
  "lead to incorrectly estimated transformation parameter values."
)

@@

@@

Here, weighted maximum likelihood estimation is based on equations \ref{eqn:box-cox-invariant-log-likelihood} and \ref{eqn:yeo-johnson-invariant-log-likelihood}.
Compared to @Raymaekers2024-zf, these log-likelihood functions includes additional terms to accommodate estimation of $x_0$ and $s$.
The weighted location- and scale-invariant Box-Cox log-likelihood function is:

```{=latex}
\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

@@

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

@@

- A step function, with $\delta_1 \geq 0$ as threshold parameter:
```{=latex}
\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
0 & \text{if } \left| \dot{x}_i \right| > \delta_1
\end{cases}
\end{equation}
```

- A triangle function (or generalised Huber weight), with $\delta_1 \geq 0$ and $\delta_2 \geq \delta_1$ as threshold parameters:
```{=latex}
\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}
```

- A tapered cosine function [@Tukey1967-eb], with $\delta_1 \geq 0$ and $\delta_2 \geq \delta_1$ as threshold parameters:
```{=latex}
\begin{equation}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation}
```

@@

Figure \ref{empirical-central-normality-kappa} shows $\tau_{\alpha = 0.95, n, \kappa}$
as a function of $n$ for different values of $\kappa$. With decreasing
$\kappa$, the test statistic curve decreases. The curves
for $\kappa = 0.60$, $\kappa = 0.70$ and $\kappa = 0.80$ are similar, whereas curves for 
$\kappa \geq 0.90$ are affected by outliers, as can be observed by comparing curves
of the central normality test with those of the empirical variant.

@@

To assess type I error rates for the (empirical) central normality test and compare
these to the Shapiro-Wilk test for normality, we randomly drew another 1000 sequences from a
standard normal distribution ($\mathcal{N}(0,1)$ for each 
$n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}$.
A second dataset was created by replacing elements with randomly drawn outliers, as described above.
For each sequence, the p-value for the respective test was used to reject the null
hypothesis that the sequence is (centrally) normal. The null hypothesis was rejected if $p \leq 0.05$.
The type I error rate was determined by computing the fraction of centrally normal sequences rejected this way.
Figure \ref{empirical-central-normality-error_rate} shows that the empirical central normality test had the expected type I error rate for
centrally normal data (with tails of the distribution not being normally distributed),
whereas the central normality test and Shapiro-Wilk test overestimated type I error rates.

@@

Outliers may be present in data and affect estimation of transformation parameters.
The log-likelihood function can be weighted to assign less weight to outlier instances,
see equations \ref{eqn:box-cox-weighted-invariant-log-likelihood} and \ref{eqn:yeo-johnson-weighted-invariant-log-likelihood}.
We proposed three weighting function: step, triangle and tapered cosine, that have one, two and two parameters, respectively.
Each weighting function then takes one of three values as input:
probabilities of the empirical distribution of the original feature,
the z-score of the transformed feature values,
or the residual error between the z-score of the transformed feature values and their expected z-score based on the normal distribution.

@@

Location- and scale-invariant transformation led to a lower overall residual errors, indicating a better transformation.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.93$).

@@

Considering single process parameters, 
the choice of learner (Friedman test: $p < 10^{-8})$),
normalisation method (Wilcoxon signed rank test: $p = 4 \cdot 10^{-8}$), 
and transformation method (Friedman test:$p = 0.007$,
all had a significant impact (at $p = 0.05$).

To estimate the marginal effects of process parameters, including transformation method, 
we first fit a regression random forest (ranger package [@Wright2017-rf] version 0.16.0): 2000 trees, 
node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), only the random forest performed better than the average of all learners (rank difference $0.173$).
The overall marginal effect of using z-standardisation to normalise features was $0.011$.
Transformation methods had the following marginal effects:
$-0.001$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

@@

In simulation, robust location- and scale-invariant transformations ranked best in datasets with outliers,
and ranked highly in datasets without outliers, except for data without outliers that were 
directly derived from strictly normal distributions.
This could be observed in real-world examples, where robust location-and scale-invariant transformations
achieved central normality for when the non-robust variant could not.

@@

Location- and scale-invariant transformations are realised by simultaneously 
optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
We derived the log-likelihood function to facilitate optimisation using MLE.
Alternatively, standardisation of a numeric variable 
(e.g., through subtracting its median value and division by its interquartile range) 
prior to conventional power transformations can achieve a similar effect in reducing sensitivity
to the distribution's location and scale, as demonstrated by our findings.
While this alternative helps prevent these issues 
-- provided that normalisation does not lead to negative values for Box-Cox transformations -- 
location- and scale-invariant transformations are able to achieve better transformations
to normality, demonstrated by a lower residual error.

@@

We introduced an empirical test for central normality to assess whether sequences
deviate from normality in a way that might require closer inspection prior to further processing.
The empirical test for central normality differs from other tests for normality, 
such as the Shapiro-Wilk test [@Shapiro1965-zd], as it assesses normality of the central portion
of a feature, instead of the entire feature.

This work has the following limitations.
Firstly, we did observe several numerical stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to very large or small numbers when using conventional power transformations.
For MLE stability issues were not observed.

>>>>

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.5, warning=FALSE}
cap <- paste0(
  "Effect of location, scale and outliers on estimation of the Box-Cox and Yeo-Johnson transformation parameter $\\lambda$. ",
  "$10000$ samples were drawn from a normal distribution: ",
  "$\\mathcal{N}(\\mu, 1)$ for the \\textit{shift} dataset, $\\mathcal{N}(10, \\sigma)$ for the \\textit{scale} dataset and $\\mathcal{N}(0, 1)$ for the \\textit{outlier} dataset. ",
  "Additionally, an outlier with value $d$ was added to the outlier dataset. ",
  "Since samples are drawn from a normal distribution, a transformation parameter of $\\lambda = 1$ is expected. ",
  "However, a large shift in location, a scale that is small compared to the location, or presence of large outliers ",
  "lead to incorrectly estimated transformation parameter values."
)

@@

First, we define a (numeric) feature. Throughout this work, whenever reference is made to a feature,
a numeric feature is meant unless noted otherwise.

\begin{definition}[Numeric feature]
A numeric feature is a finite sequence $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in  \right\}$ of length $n$.
\end{definition}

@@

Here, weighted maximum likelihood estimation is based on equations \ref{eqn:box-cox-invariant-log-likelihood} and \ref{eqn:yeo-johnson-invariant-log-likelihood}.
Compared to @Raymaekers2024-zf, these log-likelihood functions include additional terms to accommodate estimation of $x_0$ and $s$.
The weighted location- and scale-invariant Box-Cox log-likelihood function is:

\begin{definition}[Weighted log-likelihood function of the location- and scale- invariant Box-Cox power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:box-cox-invariant}).
Let $w_i \geq 0$ be the weight corresponding to each element of $\mathbf{X}$.
Let $\mu_w$ be the weighted mean of the Box-Cox transformed sequence:
\begin{equation*}
\mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation*}

Let $\sigma^2_w$ be the weighted variance of the Box-Cox transformed sequence:
\begin{equation*}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i}
\end{equation*}

Then, the weighted log-likelihood function of the location- and scale- invariant Box-Cox power transformation is:

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{definition}[Weighted log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation]

Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:yeo-johnson-invariant}).
Let $w_i \geq 0$ be the weight corresponding to each element of $\mathbf{X}$.
Let $\mu_w$ be the weighted mean of the Yeo-Johnson transformed sequence:
\begin{equation*}
\mu_w = \frac{\sum_{i=1}^n w_i \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation*}

Let $\sigma^2_w$ be the weighted variance of the Yeo-Johnson transformed sequence:
\begin{equation*}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i}
\end{equation*}

Then, the weighted log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is:

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

- A step function, with $\delta_1 \geq 0$ as threshold parameter:
\begin{equation*}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| \leq \delta_1\\
0 & \text{if } \left| \dot{x}_i \right| > \delta_1
\end{cases}
\end{equation*}

- A triangle function (or generalised Huber weight), with $\delta_1 \geq 0$ and $\delta_2 \geq \delta_1$ as threshold parameters:
\begin{equation*}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
1 - \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation*}

- A tapered cosine function [@Tukey1967-eb], with $\delta_1 \geq 0$ and $\delta_2 \geq \delta_1$ as threshold parameters:
\begin{equation*}
w_i =
\begin{cases}
1 & \text{if } \left| \dot{x}_i \right| < \delta_1\\
0.5 + 0.5 \cos\left(\pi \frac{\left| \dot{x}_i \right| - \delta_1}{\delta_2 - \delta_1} \right) & \text{if } \delta_1 \leq \left| \dot{x}_i \right| \leq \delta_2 \\
0 & \text{if } \left| \dot{x}_i \right| > \delta_2
\end{cases}
\end{equation*}

@@

Figure \ref{fig:empirical-central-normality-kappa} shows $\tau_{\alpha = 0.05, n, \kappa}$
as a function of $n$ for different values of $\kappa$. With decreasing
$\kappa$, the test statistic curve decreases. The curves
for $\kappa = 0.60$, $\kappa = 0.70$ and $\kappa = 0.80$ are similar, whereas curves for 
$\kappa \geq 0.90$ are affected by outliers, as can be observed by comparing curves
of the central normality test with those of the empirical variant.

@@

To assess type I error rates for the (empirical) central normality test and compare
these to the Shapiro-Wilk test for normality, we randomly drew another 1000 sequences from a
standard normal distribution ($\mathcal{N}(0,1)$ for each 
$n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}$.
A second dataset was created by replacing elements with randomly drawn outliers, as described above.
For each sequence, the p-value for the respective test was used to reject the null
hypothesis that the sequence is (centrally) normal. The null hypothesis was rejected if $p \leq 0.05$.
The type I error rate was determined by computing the fraction of centrally normal sequences rejected this way.
Figure \ref{fig:empirical-central-normality-error_rate} shows that the empirical central normality test had the expected type I error rate for
centrally normal data (with tails of the distribution not being normally distributed),
whereas the central normality test and Shapiro-Wilk test overestimated type I error rates.

@@

Outliers may be present in data and affect estimation of transformation parameters.
The log-likelihood function can be weighted to assign less weight to outlier instances,
see equations \ref{eqn:box-cox-weighted-invariant-log-likelihood} and \ref{eqn:yeo-johnson-weighted-invariant-log-likelihood}.
We propose three weighting functions: step, triangle and tapered cosine, that have one, two and two parameters, respectively.
Each weighting function then uses one of the following inputs:
probabilities of the empirical distribution of the original feature,
the z-score of the transformed feature values,
or the residual error between the z-score of the transformed feature values and their expected z-score based on the normal distribution.

@@

Location- and scale-invariant transformation led to a lower overall residual error, indicating a better transformation to normality.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.93$).

@@

Considering single process parameters, 
the choice of learner (Friedman test: $p < 10^{-8})$),
normalisation method (Wilcoxon signed rank test: $p = 4 \cdot 10^{-8}$), 
and transformation method (Friedman test: $p = 0.007$),
all had a significant impact (at $p = 0.05$).

To estimate the marginal effects of process parameters, including transformation method, 
we first fit a regression random forest (ranger package [@Wright2017-rf] version 0.16.0): 2000 trees, 
node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), only the random forest performed better than the average of all learners (rank difference $0.173$).
The marginal improvement in performance from z-standardisation was $0.011$.
Transformation methods had the following marginal effects:
$-0.001$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

@@

In simulation, robust location- and scale-invariant transformations ranked best in datasets with outliers,
and ranked highly in datasets without outliers, except for features without outliers that were 
directly sampled from strictly normal distributions.
In real-world examples, robust location-and scale-invariant transformations
achieved central normality when the non-robust variant could not.

@@

Location- and scale-invariant transformations are realised by simultaneously 
optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
We derived the log-likelihood function to facilitate optimisation using MLE.
Alternatively, standardisation of a numeric variable 
(e.g., through subtracting its median value and division by its interquartile range) 
prior to conventional power transformations can achieve a similar effect in reducing sensitivity
to the feature's location and scale.
While this alternative helps prevent these issues 
-- provided that standardisation does not lead to negative values for Box-Cox transformations -- 
location- and scale-invariant transformations are able to achieve better transformations
to normality, as demonstrated by lower residual errors in simulation experiments.

@@

We introduced an empirical test for central normality to assess whether sequences
deviate from normality in a way that might require closer inspection prior to further processing.
The empirical test for central normality differs from other tests for normality, 
such as the Shapiro-Wilk test [@Shapiro1965-zd], as it assesses normality of the central portion
of a feature, instead of the entire feature. 
Compared to the central normality test and the Shapiro-Wilk test, the empirical central
normality test remains consistent in the presence of outliers, although the former
tests are more powerful when outliers are absent.

This work has the following limitation: We observed several numerical stability
issues for optimisation criteria other than MLE (Appendix B). These appear in regions
where transformation parameters would lead to very large or small numbers when using conventional power transformations.
For MLE stability issues were not observed.
